---
title: "Practicum I CS5200"
author: "Pramatha Bhat(bhat.pra@northeastern.edu) & Harshitha Prabhu(prabhu.h@northeastern.edu)"
date: "Spring 2023"
output: pdf_document
---


## Connect to Database
```{r databaseConnection}
library(RMySQL)
db_user <- 'root'
db_password <- 'Appleharshi123$'
db_name <- 'MYSQL'
db_host <- 'localhost'
db_port <- 3306
mydb <-  dbConnect(MySQL(), user = db_user, password = db_password,
                 dbname = db_name, host = db_host, port = db_port)
mydb
```

## Create Database

## drop incidents table
```{sql connection=mydb, name="dropIncidentTableIfExists"}
DROP TABLE IF EXISTS incidents
```

## Create incdidents table
```{sql connection=mydb, name="createIncidentsTable"}
CREATE TABLE incidents (
    rid INTEGER PRIMARY KEY,
    `dep.date` DATE,
    origin INTEGER,
    airline INTEGER,
    aircraft TEXT,
    `flight.phase` ENUM('takeoff', 'landing', 'inflight', 'unknown'),
    altitude INTEGER CHECK (altitude >= 0),
    conditions INTEGER,
    warned BOOLEAN
);
```

## drop airports table
```{sql connection=mydb, name="dropAirportTableIfExists"}
DROP TABLE IF EXISTS airports
```

## Create airports table
```{sql connection=mydb, name="createAirportTable"}
CREATE TABLE airports (
    aid INTEGER PRIMARY KEY AUTO_INCREMENT,
    airportName TEXT,
    airportCode TEXT,
    state TEXT
);
```

## Add origin foreign key to Incidents table
```{sql connection=mydb, name = "addOriginForeignKeyToIncidentsTable"}
ALTER TABLE incidents
ADD FOREIGN KEY (origin) REFERENCES airports(aid);
```
## drop conditions table
```{sql connection=mydb, name="dropConditionsTableIfExists"}
DROP TABLE IF EXISTS conditions
```

## Create conditions table
```{sql connection=mydb, name="createConditionsTable"}
CREATE TABLE conditions (
  cid INTEGER PRIMARY KEY AUTO_INCREMENT,
  `condition` TEXT,
  explanation TEXT
);
```

## Add conditions foreign key to Incidents table
```{sql connection=mydb, name="addConditionForeignKeyToIncidentsTable"}
ALTER TABLE incidents
ADD FOREIGN KEY (conditions) REFERENCES conditions(cid);
```
## Drop airlines table
```{sql connection=mydb, name="dropAirportTableIfExists"}
DROP TABLE IF EXISTS airlines
```

## Create airlines table
```{sql connection=mydb, name="createAirlinesTable"}
CREATE TABLE airlines (
    eid INTEGER PRIMARY KEY AUTO_INCREMENT,
    airlineName TEXT,
    airlineCode TEXT,
    flag TEXT
);
```

## Add airline foreign key to Incidents table
```{sql connection=mydb, name="addAirlineForeignKeyToIncidentsTable"}
ALTER TABLE incidents
ADD FOREIGN KEY (airline) REFERENCES airlines(eid);
```
## display incident table structure
```{sql connection=mydb, eval='F', name="structureOfIncidentsTable"}
DESCRIBE incidents
```
## display airports table structure
```{sql connection=mydb, eval='F', name="structureOfAirportsTable"}
DESCRIBE airports
```
## display conditions table structure
```{sql connection=mydb, eval='F', name="structureOfConditionsTable"}
DESCRIBE conditions
```
## display airlines table structure
```{sql connection=mydb, eval='F', name="structureOfAirlinesTable"}
DESCRIBE airlines
```

## Load BirdStrikesData.csv
```{r loadCSVFile}
library(readr)
bds.raw <- read_csv("BirdStrikesData-v2.csv",show_col_types = FALSE)
bds.raw
```

## Get conditions from csv file
```{r getConditionsFromCSV}
library(RMySQL)
library(sqldf)
options(sqldf.driver = "SQLite")
conditions<-sqldf('select distinct sky_conditions FROM `bds.raw`')
conditions<-na.omit(conditions)
conditions[nrow(conditions)+1,]=c("unknown")
conditions
```

# Get airport details from the csv file BirdStrikesData.csv
```{r getAirportDetailsFromCSV}
options(sqldf.driver = "SQLite")
airport_details<-sqldf('select distinct airport, origin FROM `bds.raw`')
airport_details<-na.omit(airport_details)
airport_details[nrow(airport_details)+1,]=c("unknown")
airport_details

```

# Get airline details from the csv file BirdStrikesData.csv
```{r getAirlineDetailsFromCSV}
options(sqldf.driver = "SQLite")
airline_details<-sqldf('select distinct airline FROM `bds.raw`')
airline_details<-na.omit(airline_details)
airline_details[nrow(airline_details)+1,]=c("unknown")
airline_details

```

## get the flight details from the csv file BirdStrikesData.csv
```{r getFlightDetailsFromCSV}
options(sqldf.driver = "SQLite")
flight_details<-sqldf('select distinct flight_phase FROM `bds.raw`')
flight_details<-na.omit(flight_details)
flight_details[nrow(flight_details)+1,]=c("unknown")
flight_details

```

## get the incident details from the csv file BirdStrikesData.csv
```{r getIncidentDetailsFromCSV}
options(sqldf.driver = "SQLite")
# Read data from the raw table
incident_details <- sqldf('select distinct rid, flight_date, origin, airline, aircraft, flight_phase, altitude_ft, sky_conditions, heavy_flag FROM `bds.raw`')
for (i in 1:nrow(incident_details)) {
if(is.na(incident_details[i,3])){
incident_details[i,3] = "unknown"
 }
if(is.na(incident_details[i,4])){
 incident_details[i,4] = "unknown"
 }
}
incident_details<-na.omit(incident_details)
# Print the filtered data
incident_details
```

## insert rows from csv to airlines table
```{r insertIntoAirlines}
insert_airlines <- function() {
  
  for (i in 1:nrow(airline_details)) {
    airline<-paste0('"',airline_details[i,1],'"')
  cmd<-paste0('insert into airlines(airlineName) values (',airline,')')
      dbSendQuery(mydb,cmd)
  }
  
}
insert_airlines ()
```

## displaying 10 rows of airlines table
```{sql connection=mydb, name="displayAirlines"}
select * from airlines limit 10;
```

## insert rows into conditions table
```{r insertIntoConditions}
insert_conditions <- function() {
  
  for (i in 1:nrow(conditions)) {
    condition<-paste0('"',conditions[i,1],'"')
  cmd<-paste0('insert into conditions(`condition`) values (',condition,')')
      dbSendQuery(mydb,cmd)
  }
  
}
insert_conditions ()
```

## displaying 10 rows of conditions table
```{sql connection=mydb, name="displayConditions"}
select * from conditions limit 10;
```

## insert into airports table
```{r insertIntoAirports}
insert_airport <- function() {
  
  for (i in 1:nrow(airport_details)) {
    airport<-paste0('"',airport_details[i,1],'"')
     if(is.na(airport_details[i,2])) {
        state<-'NULL'
      } else {
      state<-paste0('"',airport_details[i,2],'"')
      }
  cmd<-paste0('insert into airports(airportName,state) values (',airport,',',state,')')
      dbSendQuery(mydb,cmd)
  }
  
}
insert_airport ()
```


## displaying 10 rows of airports table
```{sql connection=mydb, name="displayAirports"}
select * from airports limit 10;
```
## insert rows into incidents table
```{r insertIntoIncidents}
library(RMySQL)
insert_conditions <- function() {
  for (i in 1:(nrow(incident_details)-1)) {
   id<-paste0('"',incident_details[i,1],'"')
    query <- paste0('select aid from airports where state=','"',incident_details[i,3],'"')
    airport_result <- dbSendQuery(mydb, query)
    airport_data <- dbFetch(airport_result)
    airport <- airport_data[1, 1]
    date<-incident_details[i,2]
    
    if(is.na(date)) 
      {
      date<-'NULL'
      } else 
        {
        date<-strsplit(date,"/")
        date<-paste(substr(date[[1]][[3]],1,4),date[[1]][[1]],date[[1]][[2]], sep = "-")
        date<-paste0('"',date,'"')
      }
        query<-paste0('select eid from airlines where airlineName=','"',incident_details[i,4],'"')
        airlines_result <- dbSendQuery(mydb, query)
    airlines_result <- dbFetch(airlines_result)
    airline <- airlines_result[1, 1]

query<-paste0('select cid from conditions where `condition`=','"',incident_details[i,8],'"')
        conditions_result <- dbSendQuery(mydb, query)
    conditions_result <- dbFetch(conditions_result)
    condition <- conditions_result[1, 1]
    
      if(is.na(incident_details[i,5])){
         aircraft<-'NULL'
      }else
        {      
      aircraft<-paste0('"',incident_details[i,5],'"')}

    if(incident_details[i,9] == "Yes" ){
         incident_boolean = 1
      }else
        {  
          incident_boolean = 0
        }
    if(incident_details[i,6] == "Take-off run" ||  incident_details[i,6] == "Climb"){
      flight_phase = "\'takeoff\'"
    }
    else if(incident_details[i,6] == "Landing Roll" || incident_details[i,6] == "Descent"){
      flight_phase = "\'landing\'"
    }
    else if(incident_details[i,6] == "Approach" || incident_details[i,6] == "Taxi"|| incident_details[i,6] == "Parked"){
      flight_phase = "\'inflight\'"
    }
    else{
      flight_phase = "\'unknown\'"
    }
      cmd<-paste0('insert into incidents(rid, `dep.date`, origin, airline, aircraft, `flight.phase`, altitude, conditions, warned) values (',id,',',date,',',airport,',',airline,',',aircraft,',',flight_phase,',',incident_details[i,7],',',condition,',',incident_boolean,')')
    dbSendQuery(mydb, cmd)
    
  }
}
insert_conditions()
```


## display 10 rows of incidents table
```{sql connection=mydb, name="displayIncidents"}
select * from incidents limit 10;
```

## finding the 10 states with greatest number of bird strike incidents
```{sql connection=mydb, name="statesWithGreatestNumOfIncidents"} 
SELECT origin, COUNT(rid) as count FROM incidents GROUP BY origin ORDER BY count DESC LIMIT 10; 
```
## finding the airlines that have an above average number bird strike incidents.
```{sql connection=mydb, name="aboveAvgNumOfBirdStrikeIncidents"}
SELECT a.airlineName, COUNT(*) AS num_incidents
FROM incidents i
JOIN airlines a ON i.airline = a.eid
GROUP BY a.airlineName
HAVING COUNT(*) > (
    SELECT AVG(num_incidents)
    FROM (
        SELECT COUNT(*) AS num_incidents
        FROM incidents
        GROUP BY airline
    ) AS avg_incidents
)
```
## finding the number of bird strike incidents by month and by flight phase
```{r numberOfBirdStrikesByMonthAndFlightPhase}
cmd<-'SELECT distinct MONTH(`dep.date`) AS month, `flight.phase` as flight_phase, COUNT(*) AS num_incidents FROM incidents GROUP BY MONTH(`dep.date`), `flight.phase` ORDER BY month, `flight.phase`'
numOfIncidentsByMonth <- dbGetQuery(mydb, cmd)
head(numOfIncidentsByMonth, 6)
```
## Building a scatter plot that plots month along the x-axis versus number of incidents
```{r plotMonthVersusBirdIncidents}
plot(x = numOfIncidentsByMonth$month,
     y = numOfIncidentsByMonth$num_incidents,
     xlab = "Month", ylab = "No of Bird Strikes",
     main = "Month vs Number of Bird Strike Incidents",
     col = c("blue","yellow","grey"), pch=19)
     
text(numOfIncidentsByMonth$month, 
     numOfIncidentsByMonth$num_incidents, 
     labels=numOfIncidentsByMonth$num_incidents, 
     cex = 0.5, pos = 4)
legend(x="topright", legend=unique(numOfIncidentsByMonth$flight_phase),
       col = c("blue","yellow","grey"), pch=19)
```
# Drop the procedure if it already exists
```{sql connection=mydb, name="dropProcedure"}
DROP PROCEDURE IF EXISTS addNewIncident;
```


# Stored procedure to create new incident
```{sql connection=mydb, name="createAddNewIncidentProcedure"}
CREATE PROCEDURE addNewIncident (
  IN rid INTEGER,
  IN dep_date DATE, 
  IN origin INTEGER, 
  IN airline_name TEXT, 
  IN airline_code TEXT, 
  IN aircraft TEXT, 
  IN flight_phase ENUM('takeoff', 'landing', 'inflight', 'unknown'), 
  IN altitude INTEGER, 
  IN `condition` TEXT, 
  IN explanation TEXT, 
  IN airport_name TEXT, 
  IN airport_code TEXT,
  IN state TEXT,
  IN warned BOOLEAN
)
BEGIN
  DECLARE airline_id INTEGER;
  DECLARE airport_id INTEGER;
  DECLARE condition_id INTEGER;

  SELECT eid INTO airline_id FROM airlines WHERE airlineName = airline_name AND airlineCode = airline_code limit 1;

  IF airline_id IS NULL THEN
    INSERT INTO airlines (airlineName, airlineCode) VALUES (airline_name, airline_code);
    SET airline_id = LAST_INSERT_ID();
  END IF;

  SELECT aid INTO airport_id FROM airports WHERE airportCode = airport_code limit 1;

  IF airport_id IS NULL THEN
    INSERT INTO airports (airportName, airportCode, state) VALUES (airport_name, airport_code, state);
    SET airport_id = LAST_INSERT_ID();
  END IF;

  SELECT cid INTO condition_id FROM conditions WHERE `condition` = `condition` limit 1;

  IF condition_id IS NULL THEN
    INSERT INTO conditions (`condition`, explanation) VALUES (`condition`, explanation);
    SET condition_id = LAST_INSERT_ID();
  END IF;

  INSERT INTO incidents (rid, `dep.date`, origin, airline, aircraft, `flight.phase`, altitude, conditions, warned)
  VALUES (rid, dep_date, origin, airline_id, aircraft, flight_phase, altitude, condition_id, warned);
  
END
```

## Call addNewIncident Procedure with test values
```{sql connection=mydb, name="callAddNewIncidentProcedure"}
CALL addNewIncident(319594,'2022-03-03', 123, 'Delta Airlines', 'DL', 'Boeing 737', 'takeoff', 10000, 'Bird strike', 'Birds collided with engine', 'Kempegowda Vimana Nildana', 'BLR', 'Bangalore', TRUE);
```

## Testing newly added row by the procedure.
```{r testNewlyAddedRow}
result <- dbGetQuery(mydb, "SELECT * FROM incidents WHERE rid=319594")
print(result)

```


```{r disconnectDatabase}
dbDisconnect(mydb)
```